\section{Methodology Note: Spike-Guard Trigger Rate is Not a Pure Noise Meter}
\label{sec:methodology}

We added a loss-spike guard (\S\ref{sec:setup}) to the trainer to drop pathologically high-loss batches: if the current batch's average loss exceeds $1.3\times$ an EMA of recent batches, the gradient is discarded. The guard exists for a defensive reason --- to prevent a single mis-scraped pair from corrupting weights --- and we initially treated its trigger rate as a clean diagnostic of \emph{data noise}. We were wrong about that. This section documents the corrected interpretation and the data behind it.

\subsection{Trigger counts across runs}

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrr}
\toprule
\textbf{Run}                 & \textbf{Skips} & \textbf{Steps} & \textbf{per 10K steps} \\
\midrule
Base v1.1 (clean 9.3M)       & 0   & 122K & 0.00 \\
Big  v1.1 (clean 9.3M)       & 25  & 260K & 0.96 \\
Big  v2   (noisy 30M)        & 12  & 416K & 0.29 \\
\bottomrule
\end{tabular}
\caption{Spike-guard triggers across three runs, all using the same $1.3\times$EMA threshold. Naively reading these as data-noise counts gives the surprising result that Big on \emph{clean} data triggers $3.3\times$ more often than Big on \emph{noisy} data. (Base v2 was an earlier run where stdout logs were not retained; its trigger count is unrecoverable.)}
\label{tab:spike_rates}
\end{table}

If the guard rate were a pure data-noise metric, Big v1.1 (clean) and Base v2 (noisy) should differ in the obvious direction. Instead Table~\ref{tab:spike_rates} shows the \emph{opposite}: at fixed Big capacity, the clean run triggers more than the noisy run.

\subsection{The mechanism: tight-EMA false positives}

The guard threshold is $1.3\times$ \emph{the model's own EMA loss}. On clean data, Big converges to a very low loss (training floor $\sim$~2.3); the EMA tracks this floor tightly; even modest natural batch-to-batch variation can exceed $1.3\times$ a tight floor and trigger the guard. The triggered batches in this regime are essentially \emph{false positives}: the model's own well-converged baseline makes ordinary batches look spiky.

On noisy data, the loss never converges as tightly --- the noise floor stays elevated ($\sim$~3.0 for Big v2, vs.\ Big v1.1's $\sim$~2.3). The EMA is therefore looser, and only \emph{genuinely catastrophic} samples can exceed $1.3\times$ a higher floor. Triggers are rarer but each one is more likely to be real.

So the same mechanism produces \emph{more triggers on cleaner data}, because cleanliness tightens the model's loss EMA and lowers the absolute spike-detection threshold.

\subsection{What can the trigger rate be used for?}

The corrected reading: \textbf{trigger rate is a function of data noise AND the model's own convergence depth.} Two implications:

\paragraph{Same-capacity, different-data comparisons are valid.} Base v1.1 vs.\ Base v2 is a clean test of data noise (model is the same; only the data changes; differences in trigger rate isolate to data). Unfortunately the Base v2 stdout was not retained at the time of training, so we cannot report this comparison numerically. \emph{Our published v1.1 trainer now persists trigger counts to the training report so future runs are not lost.}

\paragraph{Cross-capacity, same-data comparisons are confounded.} Base v1.1 (0 triggers) and Big v1.1 (25 triggers) on identical data is \emph{not} evidence about the data --- it is evidence about the difference in convergence depth between the two models. Larger models hit lower loss floors, tightening their EMAs, and inflating their trigger counts on the \emph{same} corpus.

\paragraph{Same-capacity, same-data comparisons across runs} (e.g.\ measuring the trigger rate of a tokenizer-fix retraining vs.\ the original) are valid as a sanity check that data-pipeline changes did not introduce new pathological samples.

\subsection{Recommendation}

For papers that wish to use spike-guard or similar EMA-relative loss filters as a quantitative data-noise diagnostic, we recommend reporting the trigger rate \emph{conditional on convergence depth}, e.g.\ as a function of late-training loss EMA value. We do not do this here --- our trigger counts inform our intuition about v2 noise but are not a numeric claim. The 0-vs-12 Base-v1.1-vs-Big-v2 comparison we used in early drafts of the README is doubly confounded (different model, different data) and should be retired in favor of the four-way table above with its associated caveat.

This is a small finding and does not affect the headline 2$\times$2 claim. We include it because spike-guard or RHO-Loss-style~\cite{mindermann2022prioritized} filters are increasingly common in practice, and the false-positive-on-clean-data property is, in our experience, easy to miss until you accidentally produce a number that violates expectations.
