We report a from-scratch single-GPU (RTX 5090, 32GB) replication of the Transformer~\cite{vaswani2017attention} on WMT14 en-fr and en-de, and use it as a controlled testbed for the relationship between data quality, model capacity, and post-training. To probe the data-quality / capacity interaction we run a clean two-by-two: \{strict-filter 9.3M, loose-filter 30M\} $\times$ \{Base, Big\}. \textbf{The capacity return changes sign with data quality}: on clean data Big exceeds Base by $+0.56$ BLEU; on noisy data Big falls below Base by $-0.87$. The same $3.5\times$ extra parameters pay or cost depending only on the corpus they see, and the pattern qualitatively replicates on en-de at 4.5M scale. As a post-training addendum, we test whether quality-filtered SFT (top-1M of the noisy 30M by CometKiwi-22~\cite{rei2022cometkiwi}) recovers the lost ground; it does not. SFT degrades both Base and Big by 1.5--2.3 BLEU on newstest, with Big degrading more, because the QE-top of a multi-domain corpus is heavily UN/legislative --- high-quality but the wrong distribution for news. We argue this decomposes ``data quality'' into two independent axes --- translation correctness $\times$ alignment to the eval distribution --- and reference-free QE measures only the first. As reproducibility grounding for these claims, our pipeline obtains test BLEU 35.31 (Base, 60M) and 35.87 (Big, 209M) on WMT14 en-fr newstest2014 and 24.04 (Base) on en-de newstest2014, all averaged checkpoints with sacrebleu 13a~\cite{post2018call} --- within or above the paper's sacrebleu-equivalent band. We additionally offer a methodology note on loss-spike-guard trigger rates, which we initially treated as a noise meter but find to be a joint function of data noise and model convergence depth. All training code, configurations, four checkpoints, and stdout logs are released; the entire study reproduces in approximately 30 GPU-hours on a single 5090.
